---
title: "Age-overlap correlation tests"
output: html_document
---


Just in case we don't have our data saved from the last round:


```{r}
library(rgeos)
library(rgbif)
library(ENMTools)
library(ape)

all.worldclim <- raster::getData(name = "worldclim", download = TRUE, res = 10, var = "bio")
euro.worldclim <- crop(all.worldclim, extent(-10, 17, 39, 48))
raster.cor.matrix(env = euro.worldclim)
```

Just to keep stuff simple + running well we're going to cut down the number of variables.

```{r}
euro.worldclim <- euro.worldclim[[c(1,7,12,14)]]
raster.cor.matrix(env = euro.worldclim)
```

Load in iberolacerta.clade.Rd

```{r}
data("Iberolacerta.clade")

iberolacerta.clade

for(i in iberolacerta.clade$species){
  i$range <-  background.raster.buffer(i$presence.points, 50000, mask = euro.worldclim)
}
```


Old skool ARC

```{r}
range.aoc <- enmtools.aoc(iberolacerta.clade, nreps = 10, overlap.source = "range")
```

ARC on point data

```{r}
point.aoc <- enmtools.aoc(iberolacerta.clade, nreps = 10, overlap.source = "points")

point.aoc
```

Now we can do it with ENMs

```{r}
glm.aoc <- enmtools.aoc(iberolacerta.clade, nreps = 10, overlap.source = "glm",  env = euro.worldclim)
```


We can even do it with different metrics for ENMs

```{r}
# env.mx.aoc <- enmtools.aoc(iberolacerta.clade, nreps = 10, overlap.source = "mx",  env = euro.worldclim, metric = "env.D")
```

We're getting some issues because we don't have enough points for some of these species.  What if we drop them?

```{r}
ib.clade.small <- iberolacerta.clade
ib.clade.small$species$martinezricai <- NULL
ib.clade.small$species$horvathi <- NULL
ib.clade.small$tree <- drop.tip(ib.clade.small$tree, "martinezricai")
ib.clade.small$tree <- drop.tip(ib.clade.small$tree, "horvathi")
ib.clade.small <- check.clade(ib.clade.small)
ib.clade.small

# env.mx.aoc <- enmtools.aoc(ib.clade.small, nreps = 10, overlap.source = "mx",  env = euro.worldclim, metric = "env.D")
# env.mx.aoc
```


```{r}
env.bc.aoc <- enmtools.aoc(ib.clade.small, nreps = 10, overlap.source = "bc", env = euro.worldclim, metric = "env.D")
env.bc.aoc
```
