---
title: "Testing hypotheses for pairs of species using ENMTools"
output: html_document
---

# Building species

Just in case we don't have our data saved from the last round:


```{r warning = FALSE, message = FALSE}
library(rgbif)
library(ENMTools)

ibm <- occ_search(scientificName = "Iberolacerta monticola",  limit = 1500)
ibc <- occ_search(scientificName = "Iberolacerta cyreni",  limit = 1500)

keep.cols <- c("species", "decimalLatitude", "decimalLongitude")

ibm <- ibm$data[,keep.cols]
ibm <- as.data.frame(unique(ibm))
ibm <- ibm[complete.cases(ibm),]
colnames(ibm)  <- c("species", "lat", "lon")


ibc <- ibc$data[,keep.cols]
ibc <- as.data.frame(unique(ibc))
ibc <- ibc[complete.cases(ibc),]
colnames(ibc)  <- c("species", "lat", "lon")

all.worldclim <- raster::getData(name = "worldclim", download = TRUE, res = 10, var = "bio")
spain.worldclim <- crop(all.worldclim, extent(-10, 4, 35, 45))

plot(spain.worldclim[["bio1"]])
points(ibc[,c("lon", "lat")], pch = 16, cex = 0.5, col = "red")
points(ibm[,c("lon", "lat")], pch = 16, cex = 0.5, col = "blue")
```

Okay, let's turn both of these into enmtools.species objects

```{r warning = FALSE, message = FALSE}
monticola <- enmtools.species()
monticola$presence.points <- ibm[,c("lon", "lat")]
monticola$species.name <- "Iberolacerta monticola"
monticola$range <- background.raster.buffer(monticola$presence.points, 50000, mask = spain.worldclim)
plot(monticola)

cyreni <- enmtools.species()
cyreni$presence.points <- ibc[,c("lon", "lat")]
cyreni$species.name <- "Iberolacerta cyreni"
cyreni$range <- background.raster.buffer(cyreni$presence.points, 50000, mask = spain.worldclim)
plot(cyreni)

cyreni <- check.species(cyreni)
monticola <- check.species(monticola)
```

Now let's build a quick ENM for each species.  Since the background for cyreni is so small, we've got to restrict ourselves to only using 100 background points.

```{r warning = FALSE, message = FALSE}
monticola.glm <- enmtools.glm(monticola, spain.worldclim, f = pres ~ poly(bio1, 4) + poly(bio8, 4), nback = 100, test.prop = 0.3)

cyreni.glm <- enmtools.glm(cyreni, spain.worldclim, f = pres ~ poly(bio1, 4) + poly(bio8, 4), nback = 100, test.prop = 0.3)

cyreni.glm
monticola.glm

```

# Breadth and overlap

Okay, now let's look at breadth of species ENMs in geographic space.

```{r warning = FALSE, message = FALSE}
raster.breadth(cyreni.glm)
raster.breadth(monticola.glm)
```

How about in environment space?

```{r warning = FALSE, message = FALSE}
env.breadth(cyreni.glm, spain.worldclim)
env.breadth(monticola.glm, spain.worldclim)
```


Now let's calculate niche overlap in geographic space and environment space.

```{r warning = FALSE, message = FALSE}
raster.overlap(monticola.glm, cyreni.glm)
env.overlap(monticola.glm, cyreni.glm, spain.worldclim)
```

So the models are much more similar in geographic space than they are in environment space.  What could this mean?  DISCUSS




# Identity test

Let's do an identity/equivalency test!

```{r warning = FALSE, message = FALSE}
id.glm <- identity.test(monticola, cyreni, spain.worldclim, type = "glm", f = pres ~ poly(bio1, 4) + poly(bio8, 4), nreps = 20, nback = 100)
id.glm
```



# Background test

Let's do a background test!

```{r warning = FALSE, message = FALSE}
bg.glm <- background.test(monticola, cyreni, spain.worldclim, type = "glm", f = pres ~ poly(bio1, 4) + poly(bio8, 4), nreps = 20, nback = 100, test.type = "symmetric")
bg.glm
```





# Ecospat

```{r warning = FALSE, message = FALSE}
ib.ecospat.id <- enmtools.ecospat.id(monticola, cyreni, spain.worldclim, nreps = 100, layers = c("bio1", "bio8"))
ib.ecospat.id

ib.ecospat.bg <- enmtools.ecospat.bg(monticola, cyreni, spain.worldclim, nreps = 100, layers = c("bio1", "bio8"))
ib.ecospat.bg
```


# MOSES

```{r warning = FALSE, message = FALSE}
ib.moses <- moses.list(list(cyreni, monticola), spain.worldclim[[c(1,8)]])
ib.moses
```

Okay not much exciting there.  Let's load in an enmtools.clade object though.

```{r warning = FALSE, message = FALSE}
data(iberolacerta.clade)
iberolacerta.clade
plot(iberolacerta.clade$tree)

all.worldclim <- raster::getData(name = "worldclim", download = TRUE, res = 10, var = "bio")
euro.worldclim <- crop(all.worldclim, extent(-10, 17, 39, 48))
euro.worldclim <- euro.worldclim[[c(1,7,12,14)]]
raster.cor.matrix(env = euro.worldclim)

ib.moses.1 <- moses.list(list(iberolacerta.clade$species$aurelioi, iberolacerta.clade$species$aranica), env = euro.worldclim)

ib.moses.1

ib.moses.2 <- moses.list(list(iberolacerta.clade$species$monticola, iberolacerta.clade$species$martinezricai), env = euro.worldclim)

ib.moses.2

montmart <- combine.species(list(iberolacerta.clade$species$monticola, iberolacerta.clade$species$martinezricai))

ib.moses.3 <- moses.list(list(montmart, iberolacerta.clade$species$cyreni), env = euro.worldclim)

ib.moses.3

araur <- combine.species(list(iberolacerta.clade$species$aranica, iberolacerta.clade$species$aurelioi))
# Says to keep cyreni seprate


ib.moses.4 <- moses.list(list(araur, iberolacerta.clade$species$bonnali), env = euro.worldclim)

ib.moses.4
# Says to keep separate but not super convincing

```



# ARC and AOC

Let's start out by looking at range overlap as a function of time in Iberolacerta using raster maps:

```{r warning = FALSE, message = FALSE}
ib.arc <- enmtools.aoc(iberolacerta.clade, overlap.source = "range", nreps = 10)
ib.arc
```


And follow that up by looking at niche overlap using Bioclim  models:

```{r warning = FALSE, message = FALSE}
ib.aoc.bc <- enmtools.aoc(iberolacerta.clade, overlap.source = "bc", nreps = 10, env = euro.worldclim)
```

