---
title: "Intro to ENM and SDM"
output: html_document
---

This is going to be just a quick walk-through.  For a full tutorial, check out Jane Elith and Robert Hijmans' vignette at:

https://cran.r-project.org/web/packages/dismo/vignettes/sdm.pdf


```{r}
library(dismo)
library(rgeos)
library(rgbif)

ibm <- occ_search(scientificName = "Iberolacerta monticola",  limit = 1500)

ibm <- subset(ibm$data, select = c("species", "decimalLatitude", 
                                         "decimalLongitude"))
ibm <- as.data.frame(unique(ibm))
ibm <- ibm[complete.cases(ibm),]
colnames(ibm)  <- c("species", "lat", "lon")


all.worldclim <- raster::getData("worldclim", res = 10, var = "bio")
spain.worldclim <- crop(all.worldclim, extent(-10, 4, 35, 45))

plot(spain.worldclim[["bio1"]])
points(ibm[,c("lon", "lat")], pch = 16)
```

Okay, so we've got environmental data and occurrence data.  How do we build a model?

Some methods (e.g., bioclim or domain) can work with just the data we've got.  

```{r}
ibm.bc <- bioclim(spain.worldclim, ibm[,c("lon", "lat")])
plot(ibm.bc)
# Notice that the plot is only two variables, but our model uses 19!

response(ibm.bc)


attributes(ibm.bc)

# To make a map, we'll need to use dismo's predict function

ibm.bc.pred <- predict(object = ibm.bc, spain.worldclim)
plot(ibm.bc.pred)

points(ibm[,c("lon", "lat")], pch = 16, cex = 0.5)

# Let's stop and think about what a Bioclim model assumes about our species' response to the environment

```

Class exercise: do the same but using the domain function.  Don't worry that plot(ibm.dm) doesn't work - there's just not a plot function associated with domain models in dismo.  We'll do a much better job of visualizing these models later.  

```{r}

ibm.dm <- dismo::domain(spain.worldclim, ibm[,c("lon", "lat")])

attributes(ibm.dm)

response(ibm.dm)

# To make a map, we'll need to use dismo's predict function

ibm.dm.pred <- predict(object = ibm.dm, spain.worldclim)
plot(ibm.dm.pred)

points(ibm[,c("lon", "lat")], pch = 16, cex = 0.5)
```


We can't really evaluate the predictive capacity of our models without some absence or background data, so let's make some.  Here's one way, using circular buffers around our occurrence points.

```{r}
buffer <- circles(ibm[,c("lon", "lat")], d=100000, lonlat=TRUE)

plot(buffer)

pol <-  gUnaryUnion(buffer@polygons)

buffer.raster <- mask(spain.worldclim[[1]], pol)

plot(buffer.raster)
points(ibm[,c("lon", "lat")], pch = 16, cex = 0.5)

background <- sampleRandom(buffer.raster, size=1000, xy=TRUE)
head(background)

# EXERCISE: let's create a data frame that just has lats and lons for presence
# and another one that just has lats and longs for background data.  We'll
# give them the same column names and order, just to keep it tidy

ibm.pres <- ibm[,c("lon", "lat")]

ibm.bg <- background[,c("x", "y")]
colnames(ibm.bg) <- c("lon", "lat")

head(ibm.pres)
head(ibm.bg)

plot(ibm.dm.pred)
points(ibm.bg, pch = 4, cex = 0.5, col = "red")
points(ibm.pres, pch = 16, cex = 0.5)

# Evaluate the model using the evaluate function from dismo
ibm.dm.eval <- evaluate(p = ibm.pres, a = ibm.bg, model = ibm.dm, x = spain.worldclim)

ibm.dm.eval

attributes(ibm.dm.eval)

plot(ibm.dm.eval, "ROC")
plot(ibm.dm.eval, "TPR")
plot(ibm.dm.eval, "kappa")


# Compare to bioclim model
ibm.bc.eval <- evaluate(p = ibm.pres, a = ibm.bg, model = ibm.bc, x = spain.worldclim)

ibm.bc.eval

attributes(ibm.bc.eval)

plot(ibm.bc.eval, "ROC")
plot(ibm.bc.eval, "TPR")
plot(ibm.bc.eval, "kappa")

```

That's only training data, though!  We usually prefer to set aside some data for testing.  We'll do a quick k-fold partitioning, with k = 5

```{r}
group <- kfold(ibm.pres, 5)
group

# All this does is randomly assign a number from 1 to 5 for each presence point.
# To use these partitions for training and testing a model, we can do this:

ibm.bc <- bioclim(spain.worldclim, ibm.pres[group != 1,])

ibm.bc.pred <- predict(object = ibm.bc, spain.worldclim)
plot(ibm.bc.pred)
points(ibm.pres[group != 1,], pch = 16, cex = 0.5, col = "green")
points(ibm.pres[group == 1,], pch = 16, cex = 0.5, col = "blue")

# So now the green points are the ones used to train the model and the blue
# ones are the ones we'll use to test it

ibm.bc.eval.train <- evaluate(p = ibm.pres[group != 1,], a = ibm.bg, model = ibm.bc, x = spain.worldclim)

ibm.bc.eval.test <- evaluate(p = ibm.pres[group == 1,], a = ibm.bg, model = ibm.bc, x = spain.worldclim)


ibm.bc.eval.train
ibm.bc.eval.test

plot(ibm.bc.eval.train, "ROC")
plot(ibm.bc.eval.test, "ROC")

```

We're going to fit a model manually using one more method: GLM.  This requires our data to be in a different format.

See if you can format your data so that presence and background data are in the same table, with a new column "pres" that contains a value of 1 where the species is present and 0 where it isn't.

```{r}
pres <- c(rep(1, nrow(ibm.pres)), rep(0, nrow(ibm.bg)))
latlon <- rbind(ibm.pres, ibm.bg)
ibm.df <- cbind(latlon, pres)
head(ibm.df)
tail(ibm.df)

# We also need to get our environmental data into our data frame

pa.env <- extract(spain.worldclim, ibm.df[,1:2])
pa.env <- as.data.frame(pa.env)
pa.env$pres <- ibm.df$pres
head(pa.env)



# Here's one way to do it
ibm.glm <- glm(pres ~ ., data = pa.env)
plot(predict(spain.worldclim, ibm.glm))
points(ibm.pres, pch =16, cex = 0.5, col = "black")

# It's complaining about a rank-deficient fit, which means we have too many
# predictor variables for the number of observations we have.  Let's
# fit a simpler model.

ibm.glm <- glm(pres ~ bio1 + bio8, data = pa.env)
plot(predict(spain.worldclim, ibm.glm))
points(ibm.pres, pch =16, cex = 0.5, col = "black")

# What does our model look like?
response(ibm.glm)

# Let's fit a quadratic model instead
ibm.glm <- glm(pres ~ poly(bio1, 2) + poly(bio8, 2), data = pa.env)
ibm.glm
plot(predict(spain.worldclim, ibm.glm))
points(ibm.pres, pch =16, cex = 0.5, col = "black")
# response(ibm.glm)

# dismo does NOT like plotting responses for polynomial glms!
# We'll come back to that when we get into ENMTools

```


We can also use these models to predict to other places or time periods.

```{r}
# You can now use that model to predict habitat suitability in other places
# but note that this is SUPER slow for domain models
plot(predict(all.worldclim, ibm.glm))

```



Exercise 1: Can you figure out how to evaluate the model ibm.glm?

Exercise 2: Can you figure out how to increase the radius from which you draw your background data?  What happens to your model evaluation metrics when you do?
