---
title: "Spatial data, mapping with Leaflet"
author: "Dan Warren"
date: "1/4/2019"
output:
  html_document: default
  pdf_document: default
---

```{r}
library(raster)   
library(rgeos)
library(ggplot2)
library(viridis)
library(sp)
library(plotly)
library(leaflet)
library(maptools)
library(mapview)
library(rgbif)
```

# Retrieving spatial data and basic plots 

In this lesson, we are going to explore spatial data.  These are the three main spatial data types.  There's a lot to learn here, and we're just just going to do the bits that are going to be most crucial for the coming lessons.  For a more thorough tutorial I highly recommend the tutorials listed by Francisco Rodriguez-Sanchez here:

https://github.com/Pakillo/GISwithR/

And also Robert Hijmans' online texts at 

http://www.rspatial.org/

Although a few of the examples don't seem to be working now due to updates with the required packages (i.e., changes in syntax in newer versions).

First we're going to get a polygon for Spain.  

Oddly enough, it's the "raster" package that makes this super easy.  The getData function hooks into several online databases of spatial information, including the Global Administrative areas.

In my case, I have to call getData using raster::getData so that it knows I want the "getData" function from package "raster".  This is because another package called "nlme" also has a getData function, and R might call it by default instead.  This may not be necessary for you.

```{r}
spain <- raster::getData('GADM', country='ESP', level=1)
```


That shapefile is at a VERY fine resolution and slow to draw, because the level of detail is very, very high!  Let's simplify it.  We'll use a function called gSimplify.  In gSimplify, the "tol" parameter controls how far the resulting curve is allowed to diverge from the original.  Smaller values = more complex curves = larger and slower objects in memory.  Play around with the tol value and see what happens!

```{r}
spain <- gSimplify(spain, tol=0.01, topologyPreserve=TRUE)
```
Now let's plot it using R's base graphics.

```{r}
plot(spain)
```
This is a spatial polygon.  Polygons are collections of points that are 

Base plots leave a lot to be desired, though, so we'll use ggplot to make a map.

```{r}
ggplot() + geom_polygon(data = spain, aes(x=long, y = lat, group = group)) + 
  coord_fixed(1.3) + theme_void()
```

Adding some points

```{r}
x <- rnorm(n = 20, mean = -3, sd = 1)
y <- rnorm(n = 20, mean = 40, sd = 1)
pts <- as.data.frame(cbind(x, y))

ggplot() + geom_polygon(data = spain, aes(x=long, y = lat, group = group, alpha = 0.5)) + 
  coord_fixed(1.3) + geom_point(data = pts, aes(x = x, y = y)) +theme_void()
```


And to round it all out, we'll get some environmental data in a raster.  You can use the same getData function to get Worldclim data, which people use for SDMs and other stuff.

First we'll plot it with base graphics.

```{r}
env <- raster::getData("CMIP5", var = "tmax", res = 2.5, model = "AC", year = 70, rcp = 85)
plot(env[[1]])
```


That's the whole world, though.  We can clip that down using our Spain polygon, 

```{r}
env.poly <- crop(x = env, spain)
plot(env.poly[[1]])
```

Notice what it did there?  It didn't actually give us the portion covered by our polygons; it clipped the raster to the *extent* of that polygon layer.

If we want, we can just define an extent and use that to clip the raster.  We'll go ahead and just select the hottest month and use that from now on.  Note that this is just giving us the month with the highest high temp, not doing it by grid cell.

```{r}
spain.extent <- extent(-10, 4, 35, 45)
env <- crop(x = env, y = spain.extent)
tmax <- max(env)

plot(tmax)
points(pts)
```

And now with ggplot.  The trick to plotting raster with ggplot is that you need to convert them to data frames first.  The xy = TRUE argument tells it to record the latitude and longitude of each grid cell in the resulting data fram along with the layer value.

```{r}
tmax.df <- as.data.frame(tmax, xy = TRUE)
head(tmax.df)

ggplot() + geom_raster(data = tmax.df, aes(x = x, y = y, fill = layer)) + coord_equal() +
  geom_point(data = pts, aes(x = x, y = y)) + theme_void() + scale_fill_continuous(low="darkblue", high="red", 
                       guide="colorbar",na.value="white")
```





# Leaflet

This is where it gets really interesting.  Leaflet is a package that makes it very, very easy to make pretty, interactive maps.  It includes built-in access to some very nice online tile sets, but it's easy to throw your own data into it as well!  There's a fantastic online manual here:

(https://rstudio.github.io/leaflet/)

Now let's map the location of my favorite tapas place.

```{r}
m <- leaflet()
m <- addTiles(m)

m <- addMarkers(m, lat = 41.3818491, lng = 2.182865, label = "Can Paixano")
m
```



There's more than one way to add text information to plots, though.  Let's try it as a "popup".

```{r}

m <- leaflet()
m <- addTiles(m)
m <- addPopups(m, lat = 41.3818491, lng = 2.182865, popup = "Can Paixano")
m

```


We can manually add circles wherever we want by passing a latitude and longitude, or vectors of latitudes and longitudes.

```{r}
corners.lat <- c(41.2, 41.2, 41.4, 41.4)
corners.lng = c(2.1, 2.2, 2.2, 2.1)
addCircles(m, corners.lng, corners.lat)

```


Now let's use these to draw a rectangle.


```{r}
addPolygons(m, corners.lng, corners.lat)
```

Exercise: Can you make that into an hourglass shape?


We don't have to build polygons manually, of course; we can just plot our existing shapefiles over our leaflet maps.
```{r}
m <- leaflet()
m <- addTiles(m)
m <- addPolygons(m, data = spain, color = "black", fillColor = heat.colors(14, alpha = NULL), weight = 1, opacity = 1)
m
```

The default maps for Leaflet are nice, but you can also add tiles from other providers.  There's a list here:

(https://leaflet-extras.github.io/leaflet-providers/preview/)

```{r}
m <- leaflet()
m <- addProviderTiles(m, provider = "Stamen.Watercolor")
m
```



And you can add custom icons!
```{r}
greenLeafIcon <- makeIcon(
  iconUrl = "http://leafletjs.com/examples/custom-icons/leaf-green.png",
  iconWidth = 38, iconHeight = 95,
  iconAnchorX = 22, iconAnchorY = 94,
  shadowUrl = "http://leafletjs.com/examples/custom-icons/leaf-shadow.png",
  shadowWidth = 50, shadowHeight = 64,
  shadowAnchorX = 4, shadowAnchorY = 62
)

m <- addMarkers(m, lat = 41.3818491, lng = 2.182865, label = "Can Paixano", icon = greenLeafIcon)

m
```


One thing about leaflet is that it's really intended for interactive maps.  You can make a static map and render it to file, though, using the mapshot function in mapview.

```{r}
webshot::install_phantomjs()
mapshot(m, file = "mymap.png")
mapshot(m, file = "mymap.pdf")
```

Or, if you're in R Studio, you can plot your map in the viewer window and then use Export -> Save as Web Page to save the dynamic zoomable version to view in your web browser.

You can also add scale bars!
```{r}
m <- addScaleBar(m)
m
```

And graticules!  You may have to zoom out to see this one.
```{r}
m <- addSimpleGraticule(m, interval = 2)
m
```




You can also add rasters to leaflet maps.

```{r}
m <- leaflet()
m <- addProviderTiles(m, provider = "Stamen.Toner")
m <- addRasterImage(m, env[[8]], opacity = 0.6)
m
```

Let's do that again but define our own color ramp and add a legend.

```{r}
m <- leaflet()
m <- addProviderTiles(m, provider = "Stamen.Toner")
pal <- colorNumeric(c("#0000FF", "#FFFFCC", "#FF0000"), values(env[[8]]),
  na.color = "transparent")
m <- addRasterImage(m, colors = pal, env[[8]], opacity = 0.6)
m <- addLegend(m, pal = pal, values = values(env[[8]]),
    title = "Surface temp")
m
```

Now let's do it without the background map, but with labels for places using provider tiles!
```{r}
m <- leaflet()
pal <- colorNumeric(c("#0000FF", "#FFFFCC", "#FF0000"), values(env[[8]]),
  na.color = "transparent")
m <- addRasterImage(m, colors = pal, env[[8]], opacity = 1)
m <- addLegend(m, pal = pal, values = values(env[[8]]),
    title = "Surface temp")
m <- addProviderTiles(m, provider = "CartoDB.PositronOnlyLabels")
m
```

Now let's just play around and make something pretty.
```{r}
m <- leaflet(spain)
m <- addProviderTiles(m, provider = "NASAGIBS.ViirsEarthAtNight2012")
m <- addProviderTiles(m, provider = "OpenRailwayMap")
m
```

# Spatial data manipulation

Let's get some data from Worldclim and crop it using our extent object for Spain.  Then we'll plot it using a cool trick in the addCircleMarkers function that lets us group points that are too close together to display well.

```{r}
all.worldclim <- raster::getData("worldclim", res = 10, var = "bio")
plot(all.worldclim[["bio1"]])
spain.worldclim <- crop(all.worldclim, spain.extent)

m <- leaflet(pts)
m <- addProviderTiles(m, provider = "Stamen.Toner")
m <- addRasterImage(m, spain.worldclim[["bio1"]], opacity = 0.6)
m <- addCircleMarkers(m, lng = ~x, lat = ~y, clusterOptions = leaflet::markerClusterOptions())
m
```

While we're at it, I'll point out that you can use the %>% operator from dplyr with leaflet to make your code prettier.  Since the %>% operator automatically funnels the output from the previous function to the first argument of the input function, you can eliminate all of the "m" arguments.

```{r}
library(dplyr)
m <- leaflet(pts) %>% 
  addProviderTiles(provider = "Stamen.Toner") %>%
  addRasterImage(spain.worldclim[["bio1"]], opacity = 0.6) %>% 
  addCircleMarkers(lng = ~x, lat = ~y, clusterOptions = leaflet::markerClusterOptions())

m
```



What if we want the environmental data from our worldclim layer at each of our point locations?

```{r}
env.points <- extract(spain.worldclim, pts)
env.points

# We can also attach that to our lat/lon data

env.points <- cbind(pts, env.points)
env.points
```


Let's see how our fake species' environmental distribution differs from the suite of available habitats.  We'll need some points randomely chosen from the background.

```{r}
bg.pts <- sampleRandom(spain.worldclim, xy = TRUE, size = 1000)
head(bg.pts)
```

Notice how this got us our x and y as well as our environmental data?  Pretty slick.

```{r}
bg.pts <- as.data.frame(bg.pts)
m <- leaflet(bg.pts) %>% 
  addProviderTiles(provider = "Stamen.Toner") %>%
  addRasterImage(spain.worldclim[[1]], opacity = 0.6) %>% 
  addCircleMarkers(lng = ~x, lat = ~y, clusterOptions = leaflet::markerClusterOptions())

m
```

Okay, let's combine our presence points and our background points.  First let's make a presence/absence variable.


```{r}
pa <- c(rep(1, nrow(env.points)), rep(0, nrow(bg.pts)))

pa

```

Now we'll attach it to our data.

```{r}
pa.env.points <- rbind(env.points, bg.pts)
pa.env.points <- cbind(pa.env.points, pa)
pa.env.points <- data.frame(pa.env.points)
head(pa.env.points)
```


Let's plot where we find our species vs. the available environment.  Try visualizing several different environmental variables.

```{r}
ggplot(pa.env.points, aes(x=bio1, fill=as.factor(pa))) + geom_histogram(binwidth=2, alpha=.5, position="identity")
```


Let's try something different now - can we get a localized convex hull range map for our species?  We can, with the adehabitatHR package!

```{r}

library(adehabitatHR)

pts.locoh <- LoCoH.k(SpatialPoints(pts), k = 8)
pts.locoh.dissolved <- unionSpatialPolygons(pts.locoh, IDs = rep(1, length(pts.locoh)))
plot(spain.worldclim[["bio1"]])
plot(pts.locoh.dissolved, add = TRUE, col = "red")
points(pts, col = "blue", pch = 16)

m <- leaflet(pts.locoh.dissolved) %>% 
  addProviderTiles(provider = "Stamen.Toner") %>%
  addRasterImage(spain.worldclim[[1]], opacity = 0.6) %>% 
  addPolygons() 

m
```


Can we repeat the exercise above where we're comparing species' environmental distributions to the available environment, but use the convex hull instead of the raster for our background data?  It takes a bit more work, but we can.

```{r}
hull.pts <- spsample(pts.locoh.dissolved, n = 1000, type = "random")
class(hull.pts)
```

This hull.pts is a SpatialPoints object stores its lats and longs inside of the object in a data frame you access using @coords

```{r}
plot(spain.worldclim[[1]])
points(hull.pts@coords, pch = 16)
```

Now we'll extract our environmental data for our new background points.
```{r}
hull.env <- cbind(hull.pts@coords, extract(spain.worldclim, hull.pts@coords))
head(hull.env)
```

Then we'll repeat the steps above to visualize presence and background.  We'll just rebuild the env.points data frame since it's got the other pa data in it.
```{r}
env.points <- cbind(pts, extract(spain.worldclim, pts))
hull.pa <- c(rep(1, nrow(env.points)), rep(0, nrow(hull.env)))
hull.pa.env <- rbind(env.points, hull.env)
hull.pa.env <- cbind(hull.pa.env, hull.pa)
head(hull.pa.env)
```

And now to plot!  Note that ggplot likes data frames, not matrices, so we have to convert first.


```{r}
hull.pa.env <- data.frame(hull.pa.env)
ggplot(hull.pa.env, aes(x=bio1, fill=as.factor(pa))) + geom_histogram(binwidth=2, alpha=.5, position="identity")
```
